\textsc{HubbleCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-01-03 paper/main.tex
Draft updates

\definecolor{xlinkcolor}{rgb}{0.7752941176470588, 0.22078431372549023, 0.2262745098039215}
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

@@

\begin{abstract}
We present a multi-modal model which associates astronomical observations imaged by the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a base CLIP model using successful proposal abstracts corresponding to HST observations, summarized via guided large language (LLM) model generation. We show that the model embodies a meaningful joint representation between observations and text through experiments targeting observation retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). The model demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST) via the \texttt{astroquery} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. Note that these are not science-grade observations, but rather low-resolution images used for quick-look purposes; given the nature of associations we aim to learn, we deem this adequate for our current purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations. Images are centered and resized to a resolution per side of 512 pixels before saving. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, the abstract is excluded from the dataset.

@@

The open-weights, instruction-tuned model \textsc{Mixtral-8x7B}\footnote{\url{https://huggingface.co/mistralai/Mixtral-8x7B-Instruct-v0.1}} is used to generate the summaries, with guided generation performed using the \texttt{outlines}\footnote{\url{https://github.com/outlines-dev/outlines}} package.

@@

CLIP \citep[Contrastive Language-Image Pretraining;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss. Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$. Image and text encoders $f(\cdot)$ and $g(\cdot)$ are used to map images and text to a common embedding space. The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function, as used by CLIP, is particularly effective for multimodal learning. This bidirectionality is crucial as it ensures the model learns to map both images to text and text to images with equal importance. This symmetry in learning is essential for tasks that require a mutual understanding and interchangeability between visual and textual representations, such as image captioning, text-to-image synthesis, and cross-modal retrieval. The bidirectional loss is given by \citep{radford2021learning}
\begin{equation}
\mathcal L(\mathcal B) = -\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\underbrace{\log \frac{e^{{x}_i \cdot {y}_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{{x}_i \cdot {y}_j / \tau}}}_{\text {image } \rightarrow \text { text softmax }}+\underbrace{\log \frac{e^{{x}_i \cdot {y}_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{{x}_j \cdot {y}_i / \tau}}}_{\text {text } \rightarrow \text { image softmax }}\right)
\end{equation}
where ${x}_i=\frac{f\left(I_i\right)}{\left\|f\left(I_i\right)\right\|_2}$ and ${y}_i=\frac{g\left(T_i\right)}{\left\|g\left(T_i\right)\right\|_2}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter. 
% This bidirectional approach not only enhances the model's ability to create robust and versatile embeddings but also aligns with the intuitive understanding of how humans interact with and interpret multimodal information, further bridging the gap between artificial and natural intelligence in multimodal contexts.

@@

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized data or the raw proposal abstracts. When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder. Images are randomly cropped to the native resolution of the image encoder and randomly rotated at each training step. Given the relatively modest size of the fine-tuning set, a batch size $|\mathcal B| = 32$ is used throughout. We note that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract.

We explore three different methods of training the model on our domain dataset: \emph{(1)} Fine-tuning the entire network, starting from the pre-trained base model; \emph{(2)} Freezing the base image and text encoders, and training a small projection head on top of these; and \emph{(3)} Training the entire model from scratch.

Models were instantiated and trained using the \package{Jax} \citep{jax2018github} ecosystem. The model is trained over 20,000 steps with 5000 linear warmup steps and cosine decay using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with peak learning rate of $10^{-5}$ and weight decay $10^{-3}$. Training takes approximately 3 hours on 4 Nvidia A100 GPUs.

% Describe evals

@@

% Effect of using summarization
% Effect of using a base model vs fine-tuning from scratch

\subsection{Fine-tuned Retrieval Accuracy}
\label{sec:retrieval_acc}

\subsection{`Zero-shot' Hypothesis and Object Retrieval}
\label{sec:zero_shot}

\subsection{Text-to-Image Retrieval}
\label{sec:tti}

@@

% \begin{figure*}[!h]
% \includegraphics[width=0.95\textwidth]{example-image-a}
% \caption{Overview of dataset}
% \label{fig:dataset}
% \end{figure*}

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/val_metrics.pdf}
\caption{Retrieval accuracy}
\label{fig:retrieval_acc}
\end{figure*}

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/itt.pdf}
\caption{Retrieval accuracy}
\label{fig:itt}
\end{figure*}

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/tti_base.pdf}
\caption{Retrieval accuracy}
\label{fig:tti}
\end{figure*}

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/tti.pdf}
\caption{Retrieval accuracy}
\label{fig:tti_base}
\end{figure*}

@@

\subsubsection*{Code}

@@

This work relied on the \package{Astroquery} \citep{2019AJ....157...98G}, \package{bitsandbytes} \citep{dettmers2022llmint8}, \package{Flax} \citep{flax2020github}, \package{Jax} \citep{jax2018github}, \package{Jupyter} \citep{Kluyver2016jupyter}, \package{Matplotlib} \citep{Hunter:2007}, \package{Numpy} \citep{harris2020array}, \package{Optax} \citep{deepmind2020jax}, \package{Outlines}, \package{Pandas}, \package{Pydantic}, \package{PyTorch} \citep{paszke2019pytorch}, \package{SciPy} \citep{2020SciPy-NMeth}, \package{Transformers} \citep{wolf2019huggingface}, and \package{wandb} \citep{wandb} software packages.

@@

\subsubsection*{Acknowledgments}
This work is supported by the National Science Foundation under Cooperative Agreement PHY-2019786 (The NSF AI Institute for Artificial Intelligence and Fundamental Interactions, \url{http://iaifi.org/}). This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of High Energy Physics of U.S. Department of Energy under grant Contract Number  DE-SC0012567. YS was supported by the Research Science Institute (RSI) program at MIT. The computations in this paper were run on the FASRC Cannon cluster supported by the FAS Division of Science Research Computing Group at Harvard University.

@@

The following prompt is used to summarize the abstracts.

\begin{lstlisting}[language=Python]
@outlines.prompt
def prompt_fn(abstract):

      """<s>[INST] You are an expert astrophysicist, with broad expertise across observational and theoretical astrophysics. You are able to extract core information from astrophysical texts.

@@

@@

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/retrieval_acc.pdf}
\caption{Retrieval accuracy}
\label{fig:retrieval_acc}
\end{figure*}

>>>>

\definecolor{xlinkcolor}{rgb}{0.7752941176470588, 0.22078431372549023, 0.2262745098039215}
\definecolor{codegreen}{rgb}{0,0.4,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

@@

\begin{abstract}
We present a multi-modal model which associates astronomical observations imaged by the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a base Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts corresponding to HST observations, summarized via guided large language model (LLM) generation. We show that the model embodies a meaningful joint representation between observations and text through experiments targeting observation retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). The model demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST) via the \package{Astroquery} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. Note that these are not science-grade observations, but rather lower-resolution images used for quick-look purposes; given the nature of associations we aim to learn, we deem this adequate for our current purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations. Images are centered and resized to a resolution per side of 512 pixels before saving. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, the abstract is excluded from the dataset.

@@

The open-weights, instruction-tuned model \textsc{Mixtral-8x7B-Instruct}\footnote{\url{https://huggingface.co/mistralai/Mixtral-8x7B-Instruct-v0.1}} is used to generate the summaries, with guided generation performed using the \package{Outlines}\footnote{\url{https://github.com/outlines-dev/outlines}} package.

@@

CLIP \citep[Contrastive Language-Image Pretraining;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss. Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$. Image and text encoders $f(\cdot)$ and $g(\cdot)$ are used to map images and text to a common embedding space. The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function, as used by CLIP, is particularly effective for multimodal learning. This bidirectionality is crucial as it ensures the model learns to map both images to text and text to images with equal importance. This symmetry in learning is essential for tasks that require a mutual understanding and interchangeability between visual and textual representations, such as image captioning, text-to-image synthesis, and cross-modal retrieval. The bidirectional loss is given by \citep{radford2021learning}
\begin{equation}
\mathcal L(\mathcal B) = -\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\underbrace{\log \frac{e^{{x}_i \cdot {y}_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{{x}_i \cdot {y}_j / \tau}}}_{\text {image } \rightarrow \text { text softmax }}+\underbrace{\log \frac{e^{{x}_i \cdot {y}_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{{x}_j \cdot {y}_i / \tau}}}_{\text {text } \rightarrow \text { image softmax }}\right)
\label{eq:softmax_loss}
\end{equation}
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter. 
% This bidirectional approach not only enhances the model's ability to create robust and versatile embeddings but also aligns with the intuitive understanding of how humans interact with and interpret multimodal information, further bridging the gap between artificial and natural intelligence in multimodal contexts.

@@

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized data or the raw proposal abstracts. When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder. Images are randomly cropped to the native resolution of the image encoder and randomly rotated at each training step. Given the relatively modest size of the fine-tuning set, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were seen to susceptible to overfitting. We note that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract.

We explore three different methods of training the model on our domain dataset: \emph{(1)} Fine-tuning the entire network, starting from the pre-trained base model; \emph{(2)} Freezing the base image and text encoders, and training a small projection head on top of these; and \emph{(3)} Training the entire model from scratch. For \emph{(2)}, we use a 2-layer MLP with 1024 units and a GELU activation layer, projecting to the 512-dimensional embedding space.

Models were instantiated and trained using the \package{Jax} \citep{jax2018github} ecosystem. The model is trained over 20,000 steps with 2000 linear warmup steps 
% and cosine decay 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with peak learning rate of $10^{-5}$ and weight decay $10^{-3}$. Training takes approximately 3 hours on 4 Nvidia A100 GPUs.

The model is evaluated by tracking the loss in Eq.~\ref{eq:softmax_loss} as well as the top-$k\%$ retrieval accuracy on the held out validation set over the course of training. The retrieval accuracy is defined as the fraction of associated captions which fall within the top $k\%$ of captions by cosine similarity of the (normalized) embeddings $x_i \cdot y_j$, averaged over the images in the validation set.

@@

\paragraph*{Validation metrics}

Fig.~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy on the held out validation set over the course of training, for different variations considered. The red lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled, serving as a baseline. This baseline is seen to do on par with random expectation, unlike the others, validating the presence of a positive image-text association signal in the dataset.
%
Interestingly, the base model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$. We will therefore compare the qualitative performance of the base model with the fine-tuned model on downstream retrieval tasks.

The fiducial model with LLM-guided summarization (orange lines) is seen to perform significantly better than the model using raw abstracts as captions (purple line), validating the stronger association signal in the summarized dataset we curate. Fine-tuning a small MLP head over frozen vision and text backbones (green lines) and training from scratch (blue lines) show a non-trivial improvement in the retrieval metrics compared to the random baseline as well as base model, but with deteriorated performance compared to the fiducial set-up.

\begin{figure*}[!h]
  \includegraphics[width=0.95\textwidth]{plots/val_metrics.pdf}
  \caption{The CLIP-contrastive loss from Eq.~\ref{eq:softmax_loss} and the top-10\% retrieval accuracy computed on the validation set over the course of training. Shown for the fiducial set-up (orange), dataset using raw proposal abstracts (purple), only fine-tuning a small MLP head (green), training from scratch (blue), and a baseline trained with shuffled image-text pairs.}
  \label{fig:retrieval_acc}
  \end{figure*}

\paragraph*{Image retrieval}

Having aligned the image and text representations, we can query the validation set using natural language and show the most ``relevant'' images when ranked by cosine similarity. We show this in Figs.~\ref{fig:tti_base} and \ref{fig:tti} for the base model and fine-tuned model respectively using four simple curated queries: `globular clusters', `dwarf galaxy', `SN1987A', and `cluster lensing'.

The base model shows signs of meaningful retrieval -- it returns observations that clearly visually resemble globular clusters (first row of Fig.~\ref{fig:tti_base}), for example. Beyond this, the `dwarf galaxy' query returns a mix of galaxies and globular clusters, and it is challenging to discern meaningful associations between the other retrieved images and corresponding query.

The fine-tuned model (Fig.~\ref{fig:tti}) shows strikingly different behavior. For example, it is able to return images with processing and assembly artifacts particular to the telescope (the lines through the middle), which typically receive low similarity scores with the base model. The `dwarf galaxy' images correspond to proposals aiming to measure the kinematics of the stellar cores of dwarf galaxies. Supernova SN1987 itself can be seen as the most relevant image for the `SN1987A' query. Cluster-scale gravitational lenses are returned by the `cluster lensing' query, with lensing patterns visible in the images.

\paragraph*{Text retrieval}

Finally, we can use images from the validation set as queries and retrieve the most relevant text chunks (e.g., contained objects and use cases) from a curated list. We show the result of image-to-text retrieval in Fig.~\ref{fig:itt}, for the base as well as fine-tuned models, using the same genre of observations as for the text-to-image retrieval examples. We curate a list of possible text associations by querying the \textsc{Claude} large language model for such a list, which we show in App.~\ref{app:categories}.

The top 3 text associations are shown for each image query. The `ground truth' summarized abstract is shown in the right column. The base model is seen to return a mix of relevant and less-relevant associations. While it can often return the nature of objects imaged, we observe it to seldom return scientific phenomena (e.g., `dark matter' as successfully done by the fine-tuned model in the second row). The third row (supernova 1987A) highlights interesting behavior -- the base model erroneously attributes the object at the center of the image to a gravitational lens or protoplanetary disk, while the fine-tuned model correctly identifies it as a supernova remnant (which play a crucial role for interstellar chemistry -- another returned snippet).

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/tti_base.pdf}
\caption{Image retrieval using the base CLIP model on four curated queries.}
\label{fig:tti_base}
\end{figure*}

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/tti.pdf}
\caption{Image retrieval using the fine-tuned CLIP model on four curated queries.}
\label{fig:tti}
\end{figure*}

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/itt.pdf}
\caption{Text associations from a curated list most closely matching a given image query, for both the fine-tuned and base models. The `ground truth' LLM-summarized abstract is shown in the right column.}
\label{fig:itt}
\end{figure*}

@@

We present \textsc{HubbleCLIP}, a multi-modal fountation model that associates observations imaged by the \hubble Space Telescope with natural language in a common embedding space. The model is fine-tuned from a pre-trained CLIP model on LLM-summarized versions of \hubble proposal abstracts, leveraging a noisy signal associating text and images. We show that \textsc{Hubble CLIP} significantly outperforms the base CLIP model in quantitative metrics, such as retrieval accuracy, as well as quality of text-to-image and image-to-text retrieval. The procedure demonstrates the efficacy on fine-tuning generalist pre-trained models on small amounts of domain-specific data, in particular astronomical datasets.

@@

\paragraph*{Code and data availability}

@@

\paragraph*{Software}

This work relied on the \package{Astroquery} \citep{2019AJ....157...98G}, \package{BitsAndBytes} \citep{dettmers2022llmint8}, \package{Flax} \citep{flax2020github}, \package{Jax} \citep{jax2018github}, \package{Jupyter} \citep{Kluyver2016jupyter}, \package{Matplotlib} \citep{Hunter:2007}, \package{Numpy} \citep{harris2020array}, \package{Optax} \citep{deepmind2020jax}, \package{Outlines}, \package{Pandas} \citep{2020SciPy-NMeth}, \package{Pydantic}, \package{PyTorch} \citep{paszke2019pytorch}, \package{SciPy} \citep{2020SciPy-NMeth}, \package{Transformers} \citep{wolf2019huggingface}, and \package{Wandb} \citep{wandb} software packages.

@@

\paragraph*{Acknowledgments}

This work is supported by the National Science Foundation under Cooperative Agreement PHY-2019786 (The NSF AI Institute for Artificial Intelligence and Fundamental Interactions, \url{http://iaifi.org/}). This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of High Energy Physics of U.S. Department of Energy under grant Contract Number  DE-SC0012567. YS was supported by the Research Science Institute (RSI) program at MIT. The computations in this paper were run on the FASRC Cannon cluster supported by the FAS Division of Science Research Computing Group at Harvard University.

@@

The following prompt is used to summarize the abstracts using the \package{Outlines} package interfacing with \textsc{Mixtral-8x7B-Instruct}.

\begin{lstlisting}[language=Python]
import outlines

@outlines.prompt
def prompt_fn(abstract):
      """<s>[INST] You are an expert astrophysicist, with broad expertise across observational and theoretical astrophysics. You are able to extract core information from astrophysical texts.

@@

\section{List of Categories}
\label{app:categories}

\begin{lstlisting}[language=Python]
  ["star forming galaxies", "lyman alpha", "dust", "crowded stellar field", "core-collapse supernova", "cosmology", "gravitational lensing", "supernovae", "diffuse galaxies", "globular clusters", "stellar populations", "interstellar medium", "black holes", "dark matter", "galaxy clusters", "galaxy evolution", "galaxy formation", "quasars", "circumstellar disks", "exoplanets", "Kuiper Belt objects", "solar system objects", "cosmic web structure", "distant galaxies", "galaxy mergers", "galaxy interactions", "star formation", "stellar winds", "brown dwarfs", "white dwarfs", "nebulae", "star clusters", "galaxy archeology", "galactic structure", "active galactic nuclei", "gamma-ray bursts", "stellar nurseries", "intergalactic medium", "dark energy", "dwarf galaxies", "barred spiral galaxies", "irregular galaxies", "starburst galaxies", "low surface brightness galaxies", "ultra diffuse galaxies", "circumgalactic medium", "intracluster medium", "cosmic dust", "interstellar chemistry", "star formation histories", "initial mass function", "stellar proper motions", "binary star systems", "open clusters", "pre-main sequence stars", "protostars", "protoplanetary disks", "jets and outflows", "interstellar shocks", "planetary nebulae", "supernova remnants", "red giants", "Cepheid variables", "RR Lyrae variables", "stellar abundances", "stellar dynamics", "compact stellar remnants", "Einstein rings", "trans-Neptunian objects", "cosmic microwave background", "reionization epoch", "first stars", "first galaxies", "high-redshift quasars", "primordial black holes", "resolved binaries", "binary stars"]
\end{lstlisting}

@@

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/retrieval_acc.pdf}
\caption{Retrieval accuracy}
\label{fig:retrieval_acc_supp}
\end{figure*}
